Chemical characterization of personal filter samples from the Household Air Pollution Intervention Network (HAPIN) trial for Source Apportionment

### 2024-04-29 products/manuscript/Manuscript.qmd
Project Part 6 and final
Incorporated final edits to the project (maybe)

---
title: "Elemental characterization of personal filter samples from the Household Air Pollution Intervention Network (HAPIN) trial for Source Apportionment"
author: Erick E. Mollinedo
date: '`r format(Sys.Date(), "%B %d, %Y")`'
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/sourceapp-references.bib
csl: ../../assets/diabetologia.csl
---

@@

Household air pollution (HAP), due to the combustion of solid fuels for cooking and heating, is one of the most prominent environmental health issues in low- and middle-income countries (LMICs) [@gordon_respiratory_2014]. HAP has been listed as the second environmental health risk factor worldwide, just after outdoor air pollution. Exposure to HAP has been linked to more than 2.31 million deaths in 2019 and it is associated with adverse cardiovascular and respiratory health outcomes [@murray_global_2020; @pope_health_2006]. Particulate matter with an aerodynamic diameter less than 2.5 micrometers (PM~2.5~) is one of the major air pollutants chemically and physically composed of crustal material, black carbon (BC), organic compounds, metals and others. Toxicity from PM~2.5~ depends, among many factors, on the chemical composition of the particles which heavily depend on the sources of emissions [@clasen_design_2020; @mohammed_potential_nodate].

@@

The following study aims to produce the clean dataset from the HAPIN trial in Guatemala, needed for PMF analysis. This study also aims to conduct main and exploratory analysis of this chemical species dataset to answer some questions before the PMF modelling: 1.) What chemical species are correlated among each other from these filter samples?, 2.) What are the differences in concentrations of the chemical species if the samples are categorized by type of fuel and by study arm? 3.) What chemical species are associated with the variations in concentrations of total PM~2.5~? and 4.) What exposures can be modeled as predictors of BC concentration?

@@

@@

The concentration of the chemical species are distributed in different ranges depending on their type. For example, the highest concentrations were determined to be for Black carbon (BC), with values from 1 to 96 ug/m^3^. The remaining chemical species have lower concentrations, ranging from less than 0.1 ug/m^3^ to 10 ug/m^3^. Other noticeable patterns is that some of the species were found in higher concentrations for personal samples from participants that used biomass stoves, compared to LPG stoves (@fig-concentrations). When conducting the two-sided T-test, it was determined that all of the chemical species, except Fe, Al and Zn were significantly different between type of fuel. The differences in concentrations were also analyzed by study arm, however, given that all baseline measurements belonged to biomass fueltype observations, these results were not presented but can be observed in the Supplementary material (Figure 1).

@@

#Show the table (LPG fueltype)
pmmetrics %>% gt() %>% tab_header(title = "Table 3", subtitle = md("**Metrics summary: PM2.5 Models**")) %>%
  tab_style(style = list(cell_text(weight = "bold")),
    locations = list(cells_title(groups = "title"),
      cells_column_labels(columns = everything())))
```

@@

For the model to predict black carbon concentrations, the predictors considered in the final model were: smoking, use of mosquito coils, trash burning, use of kerosene, cooking smoked foods, crop burning, exposed to a biomass stove (or a neighboring biomass stove) and type of fuel. Use of stove, incense and a generator for corn grinding were not included due to the small variability of answers (most or all of them coded as Yes or No). When comparing the linear and GLM regression models for BC, both seem to perform similarly based on the metrics (Table 4) and the residuals and observed vs predicted values plots (Supplementary: Figures 7 and 8).

@@

#Show the table (LPG fueltype)
bcmetrics %>% gt() %>% tab_header(title = "Table 4", subtitle = md("**Metrics summary: BC Models**")) %>%
  tab_style(style = list(cell_text(weight = "bold")),
    locations = list(cells_title(groups = "title"),
      cells_column_labels(columns = everything())))
```

@@

Although the scope of this study was not to conduct the source apportionment methodology, the findings reported here are helpful for the early stages of the overall positive matrix factorization (PMF) modeling of the HAPIN Guatemala personal exposure samples. Based on the limit of detection, only 10 of the 22 chemical species (elements) were deemed important to be included in the PMF model. The model is sensitive to the mass concentrations of all species used, that is why the remaining 11 species were left out of the analyses. Some important species such as lead, copper, chromium, iodine, arsenic and others could provide a better profile of the chemical composition of total PM~2.5~. However, since these were in trace concentrations, maybe they are not as important as others, in terms of determining the sources of PM~2.5~ in this location. For the chemicals that were incorporated in the final analyis, it was observed that BC had overall the highest concentrations. This could indicate that BC is a chemical component that comes from many different sources besides just biomass burning, which is one of the most expected exposures/sources that contribute to these samples. The following element in highest concentration appeared to be potassium, which again, based on literature [@rai_real-time_2020; @pervez_impact_2012], it is an indicator of biomass burning.

When comparing the mass concentrations between samples from users that cooked using biomass fuels against cooking with LPG, it was observed that this provided a more accurate picture of the differences by the LPG intervention only. As seen in the analyses, dividing the samples in control and intervention arms produced similar results that when dividing by type of fuel. However there were higher concentration values observed for some species in the intervention samples, which is explained by the inclusion of all baseline measurements into this division. As explained in the HAPIN study description [@clasen_design_2020; @johnson_air_2020], the intervention was allocated after baseline, and during baseline it was observed that the concentrations of the main pollutants was virtually the same between both arms. In conclusion, it is considered that if looking at differences, categorizing by type of fuel, provides a clearer picture of the exposure contrasts. The correlations explored between the selected chemical species provide a glimpse of what elements could be more associated in terms of a common source. The more notables are iron, silicon and aluminum, that, based on the literature, are mostly associated by a crustal source of pollution [@zhou_chemical_2014]. Crustal means ground or dust generated particles. Other patterns are not so easily visualized, but there is a possible association between magnesium and calcium, which could be an indicator of burning of some sort of plant-based material [@zhou_chemical_2014; @rai_real-time_2020; @pervez_impact_2012]. Black carbon was moderately associated with potassium, which is a possible indicator of a biomass burning source, but this moderate to low correlation indicates that it could come from many other sources.

In terms of model selection for predicting PM~2.5~ and BC concentrations, multivariate models were deemed the most appropriate due to the presence of many predictors. In the BC prediction model, the gamma GLM model seemed more appropriate compared to the linear regression model, however this may not be the most accurate and other types of model could be a better fit. It would be interesting to fit a LASSO regression model, in order to better discriminate which of the categorical exposures could be used to explain the variations in concentrations of BC. Another explanation why this model under-performed could be to the data collection process, there were just a few records of certain exposures such as trash burning, smoking, incense, such that this could have driven the model to the wrong conclusions. This is why it is believed that modelling exposures is more accurate when considering actual exposure measurements (pollutants or chemicals), instead of relying on questionnaire based exposures that could provide biased answers. On another hand, by considering only kerosene use, exposure to a neighboring biomass stove and type of fuel as accurate predictors of BC, we could infer that the model under-performed. However, it is also important to note that these are good representations of use of biomass fuel source, which again, could be the most significant source of pollution, instead of considering other such as vehicle emissions, crustal or others.

For the PM~2.5~ models, the multivariate linear regression model seems the most appropriate as seen on the evaluation criteria (Table 3 and Supplementary material). However, it is important to point out that the moderate R-squared value implies that not all of the chemical species were included in the model to explain the total mass concentrations of PM~2.5~. Besides BC, no other organic compounds were analyzed and included in the model, and it is known that some compounds like ammonia, elemental carbon, levoglucosan and many others [@gadi_source_2019] have very important roles in particulate matter composition. It was surprising that the GLM model did not perform as good as expected, usually under-predicting extreme values. One possible way to overcome this could have been to test a multivariate linear model log-transforming the outcomes and predictors, this is due to them not following a normal distribution. In terms of the most significant species, the significant contribution of potassium and black carbon only add to the evidence that biomass burning could be the most important contributor of particulate matter in this location.

One of the limitations for this analyses is that, as pointed before, some chemical species were not included in the model. This is partly because of the source of the chemical species data, in this case it was performed only transmissometry, gravimetry and x-ray fluorescence to determine some of the species. It would be impossible to determine all the chemical species that conform the PM~2.5~ filter samples, however, adding a chromatography analysis could improve the resolution of the model by detecting additional compounds. Another goal accomplished by the analyses was to compute the final concentrations and uncertainties for all the chemical species and filter samples that will be analyzed using PMF. Currently, the PMF algorithm is licensed by the Environmental Protection Agency, under the EPA PMF 5.0 software [@norris_positive_nodate]. There was an attempt to compute the whole PMF analysis for source apportionment in R, however, this algorithm is not available publicly due to the patent. However, there is one specific packages 'pmfr' found online in a Github repository (https://github.com/skgrange/pmfr) that allows working with the outputs of the EPA PMF 5.0 software to do additional analysis and plotting.

@@

Twelve chemical species from the filter samples in rural Guatemala were found important for PMF modeling: Mg, Mn, Al, Si, S, Ca, K, Zn, Ti, Fe and BC. Of these species, all of them except Al, Zn and Fe had significantly smaller mass concentrations in LPG fuel samples compared to biomass fuel samples. BC and K are two of the most available chemical species found in the samples, which can be associated to the heavy influence of biomass burning in this location. When modelling which exposures were associated to BC concentrations, kerosene use, exposure to a neighboring biomass stove and type of fuel were significant predictors according to a GLM gamma model. However, the low diversity in answers for these categorical variables could have biased the results. On the other hand, a linear regression model was found more accurate to define total PM~2.5~ mass concentrations with all the chemical species. BC, potassium and magnesium were the most significant predictors in the PM~2.5~ model. The current study also sets the final concentration and uncertainty values needed to perform source apportionment using the PMF algorithm developed by the Environmental Protection Agency.

>>>>

---
title: "Chemical characterization of personal filter samples from the Household Air Pollution Intervention Network (HAPIN) trial for Source Apportionment"
author: Erick E. Mollinedo
date: '`r format(Sys.Date(), "%B %d, %Y")`'
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/sourceapp-references.bib
csl: ../../assets/diabetologia.csl
---

@@

Household air pollution (HAP), due to the combustion of solid fuels for cooking and heating, is one of the most prominent environmental health issues in low- and middle-income countries (LMICs) [@gordon_respiratory_2014]. HAP has been listed as the second environmental health risk factor worldwide, just after outdoor air pollution. Exposure to HAP has been linked to more than 2.31 million deaths globally in 2019, and it is more prominent in LMICs. An example of this is Guatemala with an estimate of 29.79 deaths per 100,000 inhabitants [@murray_global_2020]. HAP is associated with adverse cardiovascular and respiratory health outcomes [@pope_health_2006]. Particulate matter with an aerodynamic diameter less than 2.5 micrometers (PM~2.5~) is one of the major air pollutants chemically and physically composed of crustal material, black carbon (BC), organic compounds, metals and others. Toxicity from PM~2.5~ depends, among many factors, on the chemical composition of the particles which heavily depend on the sources of emissions [@clasen_design_2020; @mohammed_potential_nodate].

@@

The following study aims to produce the clean dataset from the HAPIN trial in Guatemala, needed for PMF analysis. This study also aims to conduct main and exploratory analysis of this chemical species dataset to answer some questions before the PMF modelling: 1.) What chemical species are correlated among each other from these filter samples?, 2.) What are the differences in concentrations of the chemical species if the samples are categorized by type of fuel?, 3.) What chemical species are associated with the variations in concentrations of total PM~2.5~? and 4.) What exposures can be modeled as predictors of BC concentration?

@@

\

@@

The concentration of the chemical species are distributed in different ranges depending on their type. For example, the highest concentrations were determined to be for Black carbon (BC), with values from 1 to 96 ug/m^3^. The remaining chemical species have lower concentrations, ranging from less than 0.1 ug/m^3^ to 10 ug/m^3^. Other noticeable patterns is that some of the species were found in higher concentrations for personal samples from participants that used biomass stoves, compared to LPG stoves (@fig-concentrations). When conducting the two-sided T-test, it was determined that all of the chemical species, except Fe, Al and Zn were significantly different between type of fuel. The differences in concentrations were also analyzed by study arm, these results can be observed in the Supplementary material (Figure 1).

@@

#Show the table (LPG fueltype)
pmmetrics %>% gt() %>% tab_header(title = md("**Metrics summary: PM2.5 Models**")) %>%
  tab_style(style = list(cell_text(weight = "bold")),
    locations = list(cells_title(groups = "title"),
      cells_column_labels(columns = everything())))
```

@@

For the model to predict BC concentrations, the predictors considered in the final model were: smoking, use of mosquito coils, trash burning, use of kerosene, cooking smoked foods, crop burning, exposed to a biomass stove (or a neighboring biomass stove) and type of fuel. Use of stove, incense and a generator for corn grinding were not included due to the small variability of answers (most or all of them coded as Yes or No). When comparing the linear and GLM regression models for BC, both seem to perform similarly based on the metrics (Table 4) and the residuals and observed vs predicted values plots (Supplementary: Figures 7 and 8).

@@

#Show the table (LPG fueltype)
bcmetrics %>% gt() %>% tab_header(title = md("**Metrics summary: BC Models**")) %>%
  tab_style(style = list(cell_text(weight = "bold")),
    locations = list(cells_title(groups = "title"),
      cells_column_labels(columns = everything())))
```

@@

Although the scope of this study was not to conduct the source apportionment methodology, the findings reported here are helpful for the early stages of the positive matrix factorization (PMF) modeling of the HAPIN Guatemala personal exposure samples. Based on the limit of detection, only 10 of the 22 chemical species (elements) were deemed important to be included in the PMF model. The model is sensitive to the mass concentrations of all species used, that is why the remaining 11 species were left out of the analyses. Some important species such as lead, copper, chromium, iodine, arsenic and others could provide a better profile of the chemical composition of total PM~2.5~. However, since these were in trace concentrations, maybe they are not as important as others, in terms of determining the sources of PM~2.5~ in this location. For the chemicals that were incorporated in the final analysis, it was observed that BC had overall the highest concentrations. This could indicate that BC is a chemical component that comes from many different sources besides just biomass burning, which is one of the most expected exposures/sources that contribute to these samples. The following element in highest concentration appeared to be potassium, which again, based on literature [@rai_real-time_2020; @pervez_impact_2012], it is an indicator of biomass burning.

As seen in the analyses, dividing the samples in control and intervention arms produced similar results that when dividing by type of fuel. The high concentrations observed for some species in the intervention samples is explained by the inclusion of all baseline measurements as part of the analyses. As explained in the HAPIN study design [@clasen_design_2020; @johnson_air_2020], the intervention was allocated after baseline, and during baseline it was observed that the concentrations of the main pollutants was virtually the same between both arms. In conclusion, it is considered that if looking at differences, categorizing by type of fuel, provides a clearer picture of the exposure contrasts. The correlations explored between the selected chemical species provide a glimpse of what elements could be more associated in terms of a common source. The more notables are iron, silicon and aluminum, that, based on the literature, are mostly associated by a crustal source of pollution [@zhou_chemical_2014]. Crustal means ground or dust generated particles. Other patterns are not so easily visualized, but there is a possible association between magnesium and calcium, which could be an indicator of burning of some sort of plant-based material [@zhou_chemical_2014; @rai_real-time_2020; @pervez_impact_2012]. BC was moderately associated with K, which is a possible indicator of a biomass burning source, but this level of correlation indicates that it could come from many other sources.

In terms of model selection for predicting PM~2.5~ and BC concentrations, multivariate models were deemed the most appropriate due to the presence of many predictors. In the BC prediction model, the gamma GLM model seemed more appropriate compared to the linear regression model, however this may not be the most accurate and other types of model could be a better fit. It would be interesting to fit a LASSO regression model, in order to better discriminate which of the categorical exposures could be used to explain the variations in concentrations of BC. Another explanation why this model under-performed could be to the data collection process, there were just a few records of certain exposures such as trash burning, smoking, incense, such that this could have driven the model to the wrong conclusions. This is why it is believed that modelling exposures is more accurate when considering actual exposure measurements (pollutants or chemicals), instead of relying on questionnaire based exposures that could bias the results. On another hand, by considering only kerosene use, exposure to a neighboring biomass stove and type of fuel as accurate predictors of BC, we could infer that the model under-performed. However, it is also important to note that these are good representations of use of biomass fuel source, which again, could be the most significant source of pollution, instead of considering other such as vehicle emissions, crustal or others.

For the PM~2.5~ models, the multivariate linear regression model seems the most appropriate as seen on the evaluation criteria (Table 3 and Supplementary material). However, it is important to point out that the moderate R-squared value implies that not all of the chemical species were included in the model to explain the total mass concentrations of PM~2.5~. Besides BC, no other organic compounds were analyzed and included in the model, and it is known that some compounds like ammonia, elemental carbon, levoglucosan and others [@gadi_source_2019] have very important roles in PM composition. It was surprising that the GLM model did not perform as good as expected, usually under-predicting extreme values. One possible way to overcome this could have been to test a multivariate linear model log-transforming the outcomes and predictors due to them not following a normal distribution. In terms of the most significant species, the significant contribution of K and BC only add to the evidence that biomass burning could be the most important contributor of particulate matter in this location. In general, the results from these models suggest that biomass burning will be a major source of PM~2.5~ pollution, something to account when analyzing the PMF results. Unfortunately, the list of categorical exposure was not wide, as to include outdoor sources such as traffic emissions or any other environmental exposures that could provide evidence of sources when conducting the future source apportionment study.

One of the limitations for this analyses is that, as pointed before, some chemical species were not included in the model. This is partly because the chemical species were determined only by transmissometry, gravimetry and x-ray fluorescence. It would be impossible to determine all the chemical species that conform the PM~2.5~ filter samples, however, adding a chromatography analysis could improve the resolution of the model by detecting additional compounds. Another goal accomplished by the analyses was to compute the final concentrations and uncertainties for all the chemical species and filter samples that will be analyzed using PMF. Currently, the PMF algorithm is licensed by the Environmental Protection Agency using the PMF 5.0 software [@norris_positive_nodate]. There was an attempt to compute the whole PMF analysis for source apportionment in R, however, this algorithm is not available publicly due to the patent. However, there is one specific packages 'pmfr' found online in a Github repository (https://github.com/skgrange/pmfr) that allows working with the outputs of the EPA PMF 5.0 software to do additional analysis and plotting.

@@

Eleven chemical species from the filter samples in rural Guatemala were found important for PMF modeling: Mg, Mn, Al, Si, S, Ca, K, Zn, Ti, Fe and BC. Of these species, all of them except Al, Zn and Fe had significantly lower mass concentrations in LPG compared to biomass fuel samples. BC and K are two of the most available chemical species found in the samples, which can be associated to the heavy influence of biomass burning in this location. When modelling which exposures were associated to BC concentrations, kerosene use, exposure to a neighboring biomass stove and type of fuel were significant predictors according to a GLM gamma model. However, the low diversity in answers for these categorical exposures could have biased the results. Also, a linear regression model was found more accurate to define total PM~2.5~ mass concentrations with all the chemical species. BC, K and Mg were the most significant predictors in the PM~2.5~ model. The current study also sets the final concentration and uncertainty values needed to perform source apportionment using the PMF algorithm developed by the EPA.
